\renewcommand{\baselinestretch}{1}

% Don't indent paragraphs.
\setlength\parindent{0em}

\newenvironment{myitemize}
{ \begin{itemize}
    \setlength{\itemsep}{2pt}
    \setlength{\parskip}{0pt}
    \setlength{\parsep}{0pt}     }
{ \end{itemize}                  }

\newenvironment{myenumerate}
{ \begin{enumerate}
    \setlength{\itemsep}{3pt}
    \setlength{\parskip}{0pt}
    \setlength{\parsep}{0pt}     }
{ \end{enumerate}                }

\noindent \LARGE{{Sefa Kılıç}\\
\vspace{1ex}
\normalsize


{1621 Bluffdale Road Apt. E} \hfill{ 443-857-3152}\\
Baltimore, MD 21244 \hfill{sefakilic@gmail.com}


\section*{\textbf{Education}}
\textit{University of Maryland Baltimore County}
\begin{myitemize}
  \item Ph.D., Biological Sciences, 2016
  \item Dissertation title: Enhancing comparative genomics of transcriptional
    regulatory networks through data collection, transfer and integration
\end{myitemize}

\textit{Middle East Technical University}
\begin{myitemize}
  \item M.Sc., Computer Science, 2011
  \item Thesis title: Clustering frequent navigation patterns from website logs
    using ontology and temporal information
\end{myitemize}

\textit{Middle East Technical University}
\begin{myitemize}
  \item B.Sc., Computer Science, 2009
  \item Senior project: YGAlib, a parallel genetic algorithm library
\end{myitemize}

\section*{\textbf{Publications}}
\begin{myenumerate}
\item The Verrucomicrobia LexA-Binding Motif: Insights into the Evolutionary
  Dynamics of the SOS Response. Ivan Erill, Susana Campoy, Sefa Kılıç, Jordi
  Barbé. \textit{Frontiers in Molecular Biosciences}. 2016

\item Assessment of Transcription Factor Binding Motif and Regulon Transfer
  Methods. Sefa Kılıç, Ivan Erill. \textit{BMC Bioinformatics}. in press

\item From data repositories to submission portals: rethinking the role of
  domain-specific databases in CollecTF. Sefa Kılıç, Dinara M Sagitova,
  Shoshannah Wolfish, Benoit Bely, Mélanie Courtot, Stacy Ciufo, Tatiana
  Tatusova, Claire O’Donovan, Marcus C Chibucos, Maria J Martin, Ivan
  Erill. \textit{Database}. 2016

\item Identification and characterization of VpsR and VpsT recognition sites in
  Vibrio cholerae. David Zamorano-Sanchez, Jiunn C. Fong, Sefa Kılıç, Ivan
  Erill and Fitnat H. Yildiz. \textit{Journal of Bacteriology}. 2015

\item CollecTF: a database of experimentally validated transcription
  factor-binding sites in Bacteria. Sefa Kılıç, Elliot R. White, Dinara
  M. Sagitova, Joseph P. Cornish and Ivan Erill. \textit{Nucleic Acids
  Research}. 2014

\item Clustering Frequent Navigation Patterns from Website Logs by Using
  Ontology and Temporal Information. Sefa Kılıç, Pınar Şenkul, İsmail Hakkı
  Toroslu, \textit{Computer and Information Sciences III}. 2013

\end{myenumerate}


\section*{\textbf{Work experience}}

\textit{University of Maryland Baltimore County} \hfill{August 2011 - May 2016}\\
Teaching assistant \hfill{Baltimore, MD}
\begin{myitemize}
\item Taught Experimental Biology Laboratory, Introduction to Bioinformatics
  and Phage Genome Analysis
\end{myitemize}

\textit{Google} \hfill{June 2015 - September 2015}\\
Software engineering intern \hfill{San Bruno, CA}
\begin{myitemize}
\item Worked on YouTube channels and playlists back-end.
\item Developed a new playlist type to make YouTube playlists playable across different countries and devices.
\item Implemented a cache staleness metric for shelves and channel configurations.
\end{myitemize}

\textit{Middle East Technical University} \hfill{September 2010 - June 2011}\\
Research assistant \hfill{Ankara, Turkey}
\begin{myitemize}
\item Intelligent Web: Using traffic analysis for self-adaptive web
\item Project funded by The Scientific and Technological Research Council of Turkey
\item Investigated whether using web site ontology and time spent on each page
  improve the prediction of next visited page.
\item Data cleaned server access logs, implemented and benchmarked the method
  predicting next page to be visited.
\end{myitemize}

\textit{Bilkent University} \hfill{September 2009 - June 2010}\\
Teaching assistant \hfill{Ankara, Turkey}
\begin{myitemize}
\item Taught Fundamentals of Computer Science and Programming Languages.
\end{myitemize}

\textit{Logo Business Solutions} \hfill{May 2008 - August 2008}\\
Software testing intern \hfill{Ankara, Turkey}
\begin{myitemize}
\item Developed unit tests with Java BeanShell for the resource planning software.
\end{myitemize}

\textit{International Business Technologies} \hfill{May 2007 - August 2007}\\
Software developer intern \hfill{Kocaeli, Turkey}
\begin{myitemize}
\item Developed Linux shell scripts for internal use.
\end{myitemize}

\section*{\textbf{Skills}}
\begin{tabular}{l l}
Programming & Python, C, C++\\
Data analysis & Pandas, NumPy, SciPy, IPython, Biopython, R, Weka, Matlab \\
Data visualization & Matplotlib, ggplot2 \\
Web development & Django, HTML, CSS, Javascript \\
Databases & MySQL, MongoDB \\
Other & GNU/Linux, Git, Emacs, RStudio, \LaTeX
\end{tabular}
